A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-05-27 paper/paper.tex
Addressing some comments by Michael.

\begin{abstract}
The \emph{crossing resolution} of a non-planar drawing of a graph is the minimum angle formed by any pair of crossing edges. Recent experiments have shown that the larger the crossing resolution is, the easier it is to read and interpret a drawing of a graph. However, maximizing the crossing resolution turns out to be an NP-hard problem in general and only heuristic algorithms are known that are mainly based on appropriately adjusting force-directed algorithms.

In this paper, we propose a new randomization-based heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known ones from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with significantly better crossing resolution, but this comes at the cost of slightly higher running time, especially when the input graph is large and dense. 
\end{abstract}

@@

In Graph Drawing, there exists a really rich literature and a wide range of techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restricted), is a difficult and challenging task, mainly due to the edge crossings which negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the enstablished techniques are significantly fewer (e.g., crossing minimization heuristics~\cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview refer to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

@@

In general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems to be difficult to compute the expected quality of the produced drawings; a notable exception is the randomized approach by Goldschmidt and Takvorian~\cite{DBLP:journals/networks/GoldschmidtT94} for computing large planar subgraphs. Since we could not provide any theoretical guarantee on the expected quality of the produced drawings as well (mainly due to the nature of the problem itself), we decided to follow a more practical approach. To this end, we implemented our algorithm and the ones presented in~\cite{DBLP:journals/vlc/HuangEHL13} and~\cite{DBLP:journals/cj/ArgyriouBS13}, and we experimentally compared the crossing resolutions of the drawings produced by them for standard benchmark graphs (such as the Rome and the North graphs) that are widely used in Graph Drawing for comparing different drawing algorithms. Our experimental evaluation indicates that our new method significantly outperforms the ones that are based on adjustments of force-directed algorithms~\cite{DBLP:journals/vlc/HuangEHL13,DBLP:journals/cj/ArgyriouBS13} in terms of crossing resolution, but this comes at the cost of slightly worse running time, especially for large and dense graphs.

@@

As already mentioned, the study of the crossing resolution maximization problem has mainly focused on (properties of) RAC graphs, that is, on the optimal case of the crossing resolution maximization problem. The study was initiated by Didimo et al.~\cite{DBLP:journals/tcs/DidimoEL11}, who showed that an $n$-vertex RAC graph has at most $4n-10$ edges. 
Angelini et al.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11} presented results for two interesting variants, in which the input graph is either an acyclic digraph or a graph of bounded vertex degree. Argyriou et al.~\cite{DBLP:journals/jgaa/ArgyriouBS12} showed that the problem of deciding whether a graph is RAC is NP-hard. Didimo et al.~\cite{DBLP:journals/ipl/DidimoEL10} completely characterized the complete bipartite RAC graphs. Di~Giacomo et al.~\cite{DBLP:journals/algorithmica/GiacomoDEL14}, and Hong and Nagamochi~\cite{DBLP:conf/wg/HongN15} studied variants, in which the vertices of the input graph are restricted on two parallel lines and on the circumference of a circle, respectively. Eades and Liotta~\cite{DBLP:journals/dam/EadesL13} showed that the optimal RAC graphs (i.e., those of maximum density) are 1-planar (i.e., they admit drawings in which each edge is crossed at most once). Deciding, however, whether a 1-planar graph is RAC is NP-hard~\cite{DBLP:journals/tcs/BekosDLMM17}. Other relationships between the class of RAC graphs and subclasses of 1-planar graphs have also been studied by Bachmaier et al.~\cite{DBLP:journals/dam/BachmaierBHNR17} and Brandenburg et al.~\cite{DBLP:journals/tcs/BrandenburgDEKL16}. Note that the problem of finding RAC drawings has been also studied in the presence of bends; see e.g.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/comgeo/ArikushiFKMT12,DBLP:journals/tcs/DidimoEL11,DBLP:journals/mst/GiacomoDLM11}.

@@

In this section, we describe our heuristic algorithm for obtaining drawings with high crossing resolution. We assume that the input of our algorithm consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ that is of some crossing resolution $c(\Gamma_0)$. We also assume that no two edges of $G$ overlap in $\Gamma_0$, that is, $c(\Gamma_0)>0$. Note that a drawing $\Gamma_0$ of $G$ meeting these preconditions can be, e.g., a circular drawing of $G$ or a drawing obtained by applying a standard spring embedding algorithm on $G$.

Our algorithm is iterative and at each iteration step performs some operations that are mainly based on randomization. More precisely, at the $i$-th iteration step, we assume that we have computed a drawing $\Gamma_{i-1}$ of some crossing resolution $c(\Gamma_{i-1}) \geq c(\Gamma_0)$, where $\Gamma_0$ is our initial drawing. In other words, we assume, as an invariant property of our algorithm, that the crossing resolution cannot be decreased at some iteration step. Then, a vertex of $\Gamma_{i-1}$ is chosen arbitrary at random from a so-called \emph{vertex-pool}, which may contain:

@@

To formally define the critical vertices, we first need to introduce the notion of critical edge-pairs. A pair of edges, say $e$ and $e'$, is called \emph{critical} in drawing $\Gamma_{i-1}$, if $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is either equal to the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$. The set of critical vertices of drawing $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair.

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. As a result, their role in our algorithm is central\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations in order to converge to a good solution, because it is simply very unlike to select one of them to move.}. The reason is that by slightly changing the location of a critical vertex appropriately, one naturally expects to improve the crossing resolution of the current drawing. In fact, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing only at the critical vertices of the drawing, it is highly possible that the algorithm will get trapped to some local maxima after a number of iterations. Hence, special care is needed in order to avoid these bottlenecks, especially when the graph to be drawn is large. We discuss heuristics to avoid such bottlenecks in  Section~\ref{ssec:maxima}.

@@

\begin{itemize}
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices. 
\item If during the last $\zeta$ iterations the crossing resolution has not been improved, then the vertex-pool becomes \emph{wider} containing all the vertices of the graph, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations. 
\item After these $\zeta'$ iterations, the vertex-pool switches back to the critical vertices.
\end{itemize}
%
From our extensive practical analysis, we quickly realized that, while this approach turned out to be quite effective for medium-size graphs, for larger and denser graph, unfortunately, it was not so efficient. What we observed is that in larger graphs it was unlikely to change the embedding of the graph in a beneficial way for the algorithm to proceed, mainly because of the random choice of the vertices to be moved; in most iterations with the wider vertex-pool, the vertices chosen to be moved were not in the part of the drawing where the critical vertices resided.

>>>>

\begin{abstract}
The \emph{crossing resolution} of a non-planar drawing of a graph is the value of the minimum angle formed by any pair of crossing edges. Recent experiments have shown that the larger the crossing resolution is, the easier it is to read and interpret a drawing of a graph. However, maximizing the crossing resolution turns out to be an NP-hard problem in general and only heuristic algorithms are known that are mainly based on appropriately adjusting force-directed algorithms.

In this paper, we propose a new randomization-based heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known approaches from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with significantly better crossing resolution, but this comes at the cost of slightly higher running time, especially when the input graph is large and dense. 
\end{abstract}

@@

In Graph Drawing, there exists a really rich literature and a wide range of techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restriction), is a difficult and challenging task, mainly due to the edge crossings which negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the established techniques are significantly fewer (e.g., crossing minimization heuristics~\cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview refer to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

@@

In general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems difficult to even speculate about the expected quality of the produced drawings; a notable exception is the randomized approach by Goldschmidt and Takvorian~\cite{DBLP:journals/networks/GoldschmidtT94} for computing large planar subgraphs. Since we could not provide any theoretical guarantee on the expected quality of the produced drawings as well (mainly due to the nature of the problem itself), we decided to follow a more practical approach. To this end, we implemented our algorithm and the ones presented in~\cite{DBLP:journals/vlc/HuangEHL13} and~\cite{DBLP:journals/cj/ArgyriouBS13}, and we experimentally compared the crossing resolutions of the drawings produced by them for standard benchmark graphs (such as the Rome and the North graphs) that are widely used in Graph Drawing for comparing different drawing algorithms. Our experimental evaluation indicates that our new method significantly outperforms the ones that are based on adjustments of force-directed algorithms~\cite{DBLP:journals/vlc/HuangEHL13,DBLP:journals/cj/ArgyriouBS13} in terms of crossing resolution, but this comes at the cost of slightly worse running time, especially for large and dense graphs.

@@

As already mentioned, the study of the crossing resolution maximization problem has mainly focused on (properties of) RAC graphs, that is, on the optimal case of the crossing resolution maximization problem. The study was initiated by Didimo et al.~\cite{DBLP:journals/tcs/DidimoEL11}, who showed that an $n$-vertex RAC graph has at most $4n-10$ edges. 
Angelini et al.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11} presented results for two interesting variants, in which the input graph is either an acyclic digraph or a graph of bounded vertex degree. Argyriou et al.~\cite{DBLP:journals/jgaa/ArgyriouBS12} showed that the problem of deciding whether a graph is RAC is NP-hard. Angelini et al.~\cite{DBLP:conf/s-egc/AngeliniBDFHKLL11} observed that planar graphs of bounded degree admit straight-line RAC drawings in subquadratic area.
%Didimo et al.~\cite{DBLP:journals/ipl/DidimoEL10} completely characterized the complete bipartite RAC graphs.
Di~Giacomo et al.~\cite{DBLP:journals/algorithmica/GiacomoDEL14}, and Hong and Nagamochi~\cite{DBLP:conf/wg/HongN15} studied variants, in which the vertices of the input graph are restricted on two parallel lines and on the circumference of a circle, respectively. Eades and Liotta~\cite{DBLP:journals/dam/EadesL13} showed that the optimal RAC graphs (i.e., those of maximum density) are 1-planar (i.e., they admit drawings in which each edge is crossed at most once). Deciding, however, whether a 1-planar graph is RAC is NP-hard~\cite{DBLP:journals/tcs/BekosDLMM17}. Other relationships between the class of RAC graphs and subclasses of 1-planar graphs have also been studied by Bachmaier et al.~\cite{DBLP:journals/dam/BachmaierBHNR17} and Brandenburg et al.~\cite{DBLP:journals/tcs/BrandenburgDEKL16}. Note that the problem of finding RAC drawings has been also studied in the presence of bends; see e.g.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/comgeo/ArikushiFKMT12,DBLP:journals/tcs/DidimoEL11,DBLP:journals/mst/GiacomoDLM11}.

@@

In this section, we describe our heuristic algorithm for obtaining drawings with high crossing resolution. We assume that the input of our algorithm consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ that is of some crossing resolution $c(\Gamma_0)$. We also assume that no two edges of $G$ overlap in $\Gamma_0$, that is, $c(\Gamma_0)>0$. A circular drawing of $G$ or a drawing obtained by applying a standard force-directed algorithm on $G$ clearly meets these preconditions.

Our algorithm is iterative and at each iteration step performs some operations that are mainly based on randomization. More precisely, at the $i$-th iteration step, we assume that we have computed a drawing $\Gamma_{i-1}$ of some crossing resolution $c(\Gamma_{i-1}) \geq c(\Gamma_0)$, where $\Gamma_0$ is our initial drawing. In other words, we assume, as an invariant property of our algorithm, that the crossing resolution cannot be decreased at some iteration step. Then, a vertex of $\Gamma_{i-1}$ is chosen arbitrarily at random from a so-called \emph{vertex-pool}, which may contain:

@@

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. To formally define them, we first need to introduce the notion of critical edge-pairs. A pair of edges, say $e$ and $e'$, is called \emph{critical} in drawing $\Gamma_{i-1}$, if $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is equal to the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$. The set of critical vertices of drawing $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair.

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations in order to converge to a good solution, because it is simply very unlike to select one of them to move.}. The reason is that by slightly changing the location of a critical vertex appropriately, one naturally expects to improve the crossing resolution of the current drawing. In fact, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing only at the critical vertices of the drawing, it is highly possible that the algorithm will get trapped to some local maxima after a number of iterations. Hence, special care is needed in order to avoid these bottlenecks, especially when the graph to be drawn is large. We discuss heuristics to avoid such bottlenecks in  Section~\ref{ssec:maxima}.

@@

\begin{itemize}
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices. 
\item If during the last $\zeta$ iterations the crossing resolution has not been improved, then the vertex-pool becomes \emph{wider} containing all the vertices of the graph, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations. 
\item After these $\zeta'$ iterations, the vertex-pool switches back to the critical vertices.
\end{itemize}
%
From our extensive practical analysis, we quickly realized that, while this approach turned out to be quite effective for medium-size graphs, for larger and denser graphs, unfortunately, it was not so efficient. What we observed is that in larger graphs it was unlikely to change the embedding of the graph in a beneficial way for the algorithm to proceed, mainly because of the random choice of the vertices to be moved; in most iterations with the wider vertex-pool, the vertices chosen to be moved were not in the part of the drawing where the critical vertices resided.
